|

### 2021-07-27 prova_knit_finale.Rmd
small changes

Let us consider a more general state-space model, where the state vector includes a time-constant parameter vector $\psi \in \Psi$, where $\Psi$ is the parameter space. Since we can interpret the parameter as a state with the law of motion $\psi_t=\psi_{t-1}=\psi$, then it is possible to adopt a PF algorithm to estimate $\psi$.\
However, being $\psi$ constant over time, it is meaningless to sample its values sequentially: the first extraction of $\psi$ at the initial stage 0 (i.e., $\psi_0^n$) generates a constant path of particles for the parameter ($\psi_t^n=\psi_0^n$). \footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\
Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution. In this way, at every time, the support of $\psi$ is not limited to the initially sampled $N$ values. \
Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter. Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\
Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
The LWF can be described by the following algorithm.
\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.

@@

The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$. While it would be more intuitively immediate to sample new parameters from a Normal density centered at the previously sampled values (i.e., $\psi^n$), this would increase the variance of $\psi$ over time.\

@@

\subsection{Implementation}
\
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. The algorithm follows the following steps.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
\begin{align*}
m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}

>>>>

Let us consider a more general state-space model, where the state vector includes a time-constant parameter vector $\psi \in \Psi$, where $\Psi$ is the parameter space. Since we can interpret the parameter as a state with the law of motion $\psi_t=\psi_{t-1}=\psi$, then it is possible to adopt a PF algorithm to estimate $\psi$.\
However, being $\psi$ constant over time, it is meaningless to sample its values sequentially: the first extraction of $\psi$ at the initial stage 0 (i.e., $\psi_0^n$) generates a constant path of particles for the parameter ($\psi_t^n=\psi_0^n$). \footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\
Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution. In this way, at every time, the support of $\psi$ is not limited to the initially sampled $N$ values. \
Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter. Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF. For instance, in our implementation with the random walk model of section \ref{implementation}, we use an auxiliary particle LWF.}\
Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
The LWF can be described by the following algorithm.
\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.

@@

The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal\footnote{More generally, distributions whose support coincides with the relevant parameter space.} centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$. While it would be more intuitively immediate to sample new parameters from a Normal density centered at the previously sampled values (i.e., $\psi^n$), this would increase the variance of $\psi$ over time.\

@@

\subsection{Implementation}
\
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values. The algorithm follows the following steps.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
\begin{align*}
m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}
